Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-08-24 Paper/Paper.tex
last changes in the text, added a reference

Persons affected from physical and mental restrictions and their care givers can profit significantly from unobtrusive activity monitoring solutions. For example, formal care givers could equip a person with a wearable activity monitoring solution that evaluates the course of a disease or the influence of an adapted medication. This scenario is especially relevant for people suffering from dementia who have limitations in organizing their daily activities. A simple wearable activity monitoring solution could analyze activities performed in daily life, such as drinking, eating and sleeping habits.

@@

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{Images/hardware_sensor_2.jpg}
	\caption{The inertial data logger featuring a low-power microcontroller, a 3 axis accelerometer, a microSD flash card for storing the sensor data and a USB connector for accessing the data (on the right) is powered by a small lithium polymer battery and is packaged into a plastic case to be worn at the wrist (a version with an OLED display).}
	\label{fig:sensornode}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{Images/schematic.pdf}
	\caption{The sensing circuit is based on a timer with an operational amplifier that acts as a voltage follower. The wire that is labeled with ``antenna'' leads to the sensing electrode, whereas the wire labeled with ``guard'' leads to the shield electrode.}
	\label{fig:schematic}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_2.pdf}
	\caption{Example of the ``sitting'' activity in which the user moved his hands quite frequently (bottom plot). Most of the time the values of the proximity sensor stay more or less constant, probably due to the hands position on the couch's fabric. The sharp peak in the capacitive sensor data (upper plot) occurred when the participant scratched the back of his head. }
	\label{fig:sitting}
\end{figure}

@@

In the following, the participant were instructed to lie  down on the couch. Again, there were great variations in how this activity was performed by the participants. For example, some of them crossed their hands under their head, or placed them on their body. For this class, the average F-measure could only be increased by 2.4\%, from 81.2 to 83.6\%. Considering some participants, the activity was often confused with the ``make sandwich'' class. By using the proximity modality, the confusion between the two classes could be reduced.

\begin{figure}[t]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/tobias_4.pdf}
	\caption{An example of the ``getting things'' activity, where the participants had to get food and dishes from shelves and lockers. The proximity sensor peaks in the beginning (9s and 19s) indicate immediate proximity to shelf, and to the locker (55-63s) in the kitchen (upper plot). The signal drop at the end results from the participant placing his hand on the table when she was finished.}
	\label{fig:get_things}
\end{figure}

After that, the participants were asked to walk over to the kitchen and to put food and dishes from a shelf and a locker on the table. This activity involved direct interactions with various objects as well as proximity to furniture in the room (see Table \ref{tab:activities}). The capacitive proximity sensor was able to capture the proximity to the shelf and to the table. The average F-measure for this activity is rather low, but improved by 6.8\% from 53.8\% to 60.6\%. 
The worst performing participants for this activity reached an F-measure of 46.8\% without and 53.5\% with the capacitive sensor, while the best performing one reached 62.0\% and 64.5\% respectively. The low performance results from confusions with other activities, with a higher tendency to the ``make sandwich'' activity across all participants. This is most likely due to various objects involved in both activities, and the fact that 1-second features are obtained. The capacitive sensor modality has a much more positive impact reducing the confusion with other activities.

\begin{figure}[t]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``make sandwich'' activity, where the participants had to put marmalade on a slice of bread. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows recurring hand motions.}
	\label{fig:prep_bread}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An example of a participant eating a marmalade sandwich, taking 5 bites from it. After each bite, the hand is placed on the table, which can be recognized both in the acceleration as well as the proximity data plots.}
	\label{fig:eating}
\end{figure}

When considering the eating activity, the impact of the new capacitive proximity sensor on the classification performance is quite low, as the chosen features are able to distinguish it from other activities. Some of the participants ate their sandwich leaving their hand close to the mouth, while others moved their hand up and down putting their sandwich aside on the plate, which also results in the performance range from 71.2 to 88.1\% without and 77.5 to 90.6\% with the proximity data. An example of an eating activity is shown in Figure \ref{fig:eating}, where the participant took a few bites from the sandwich while putting it down every time. The average F-measure for that activity increased slightly by 2.7\% (from 79.5\% to 82.2\%).

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/marko_8.pdf}
	\caption{An exemplary instance of the class ``walking''. The acceleration sensor and the proximity sensor show periodic recurring patterns that are related to the pendulum-like arm movement and the proximity to the person's body during those movements.}
	\label{fig:walking}
\end{figure}

@@

\begin{figure}[t]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_9.pdf}
	\caption{During the sleeping activity the data from both sensors remains constant for large time spans. One can draw conclusions about the coverage of the arm with either cushions, blankets or the proximity to the mattress, the head or body of the participant.}
	\label{fig:sleeping}
\end{figure}

@@

\section*{Acknowledgments}

@@

\bibliographystyle{splncs03}
%\bibliography{references}

>>>>

Persons affected from physical and mental restrictions and their care givers can profit significantly from unobtrusive activity monitoring solutions. For example, formal care givers could equip a person with a wearable activity monitoring system that evaluates the course of a disease or the influence of an adapted medication \cite{Muhlsteff2004a}. This scenario is especially relevant for people suffering from dementia who have limitations in organizing their daily activities. A simple wearable activity monitoring solution could analyze activities performed in daily life, such as drinking, eating and sleeping habits.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{Images/hardware_sensor_2.jpg}
	\caption{The inertial data logger featuring a low-power microcontroller, a 3 axis accelerometer, a microSD flash card for storing the sensor data and a USB connector for accessing the data (on the right) is powered by a small lithium polymer battery and is packaged into a plastic case to be worn at the wrist (a version with an OLED display).}
	\label{fig:sensornode}
\end{figure}

@@

@@

\begin{figure}[t]
	\centering
		\includegraphics[width=1.00\textwidth]{Images/schematic.pdf}
	\caption{The sensing circuit is based on a timer with an operational amplifier that acts as a voltage follower. The wire that is labeled with ``antenna'' leads to the sensing electrode, whereas the wire labeled with ``guard'' leads to the shield electrode.}
	\label{fig:schematic}
\end{figure}

@@

\begin{figure}[h]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_2.pdf}
	\caption{Example of the ``sitting'' activity in which the user moved his hands quite frequently (bottom plot). Most of the time the values of the proximity sensor stay more or less constant, probably due to the hands position on the couch's fabric. The sharp peak in the capacitive sensor data (upper plot) occurred when the participant scratched the back of his head. }
	\label{fig:sitting}
\end{figure}

@@

In the following, the participants were instructed to lie  down on the couch. Again, there were great variations in how this activity was performed by the participants. For example, some of them crossed their hands under their head, or placed them on their body. For this class, the average F-measure could only be increased by 2.4\%, from 81.2 to 83.6\%. Considering some participants, the activity was often confused with the ``make sandwich'' class. By using the proximity modality, the confusion between the two classes could be reduced.

\begin{figure}[!t]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/tobias_4.pdf}
	\caption{An example of the ``getting things'' activity, where the participants had to get food and dishes from shelves and lockers. The proximity sensor peaks in the beginning (9s and 19s) indicate immediate proximity to shelf, and to the locker (55-63s) in the kitchen (upper plot). The signal drop at the end results from the participant placing his hand on the table when she was finished.}
	\label{fig:get_things}
\end{figure}

After that, the participants were asked to walk over to the kitchen and to put food and dishes from a shelf and a locker on the table. This activity involved direct interactions with various objects as well as proximity to furniture in the room (see Table \ref{tab:activities}). The capacitive proximity sensor was able to capture the proximity to the shelf and to the table (see Figure \ref{fig:get_things}). The average F-measure for this activity is rather low, but improved by 6.8\% from 53.8\% to 60.6\%. 
The worst performing participants for this activity reached an F-measure of 46.8\% without and 53.5\% with the capacitive sensor, while the best performing one reached 62.0\% and 64.5\% respectively. The low performance results from confusions with other activities, with a higher tendency to the ``make sandwich'' activity across all participants. This is most likely due to various objects involved in both activities, and the fact that 1-second features are obtained. The capacitive sensor modality has a more positive impact reducing the confusion with other activities.

\begin{figure}[h]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``make sandwich'' activity, where the participants had to put marmalade on a slice of bread. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows recurring hand motions.}
	\label{fig:prep_bread}
\end{figure}

@@

\begin{figure}[h]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An example of a participant eating a marmalade sandwich, taking 5 bites from it. After each bite, the hand is placed on the table, which can be recognized both in the acceleration as well as the proximity data plots.}
	\label{fig:eating}
\end{figure}

When considering the eating activity, the impact of the new capacitive proximity sensor on the classification performance is quite low, as the chosen features are able to distinguish it from other activities. Some of the participants ate their sandwich leaving their hand close to the mouth, while others moved their hand up and down putting their sandwich aside on the plate, which also results in the performance range from 71.2 to 88.1\% without and 77.5 to 90.6\% with the proximity data. An example of an eating activity is shown in Figure \ref{fig:eating}, where the participant took a few bites from the sandwich while putting it down every time. The average F-measure increased slightly by 2.7\% (from 79.5\% to 82.2\%).

@@

\begin{figure}[t]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/marko_8.pdf}
	\caption{An exemplary instance of the class ``walking''. The acceleration sensor and the proximity sensor show periodic recurring patterns that are related to the pendulum-like arm movement and the proximity to the person's body during those movements.}
	\label{fig:walking}
\end{figure}

@@

\begin{figure}[t]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_9.pdf}
	\caption{During the sleeping activity the data from both sensors remains constant for large time spans. The capacitive sensor shows the coverage of the arm with either cushions, blankets or the proximity to the mattress, the head or body of the participant.}
	\label{fig:sleeping}
\end{figure}

@@

\section*{Acknowledgements}

@@

\bibliographystyle{splncs03}
\bibliography{references}
